### 2008-12-18 talk/ecoop2009/rainbow.tex
lots of fixes

In the sequel, we describe in more details one of the main new
techniques introduced in our approach, which we call \emph{promotion}.  In
short, it allows an arbitrary run-time value to be turned into a
compile-time value at any point in time.  Promotion is thus the central way by
which we make use of the fact that the JIT is running interleaved with actual
program execution. Each promotion point is explicitly defined with a hint that
must be put in the source code of the interpreter.

@@

Promotion requires interleaving compile-time and run-time phases,
otherwise the compiler can only use information that is known ahead of
time. It is impossible in the "classical" approaches to partial
evaluation, in which the compiler always runs fully ahead of execution
This is a problem in many large use cases.  For example, in an
interpreter for a dynamic language, there is mostly no information
that can be clearly and statically used by the compiler before any
code has run.

A very different point of view on promotion is as a generalization of
techniques that already exist in dynamic compilers as found in modern
object-oriented language virtual machines.  In this context feedback
techniques are crucial for good results.  The main goal is to
optimize and reduce the overhead of dynamic dispatching and indirect
invocation.  This is achieved with variations on the technique of
polymorphic inline caches \cite{XXX}: the dynamic lookups are cached and
the corresponding generated machine code contains chains of
compare-and-jump instructions which are modified at run-time.  These
techniques also allow the gathering of information to direct inlining for even
better optimization results.

In the presence of promotion, dispatch optimization can usually be
reframed as a partial evaluation task.  Indeed, if the type of the
object being dispatched to is known at compile-time, the lookup can be
folded, and only a (possibly inlined) direct call remains in the
generated code.  In the case where the type of the object is not known
at compile-time, it can first be read at run-time out of the object and
promoted to compile-time.  As we will see in the sequel, this produces
very similar machine code \footnote{This can also be seen as a generalization of
a partial evaluation transformation called "The Trick" (see e.g. \cite{XXX}),
which again produces similar code but which is only applicable for finite sets
of values.}.

The essential advantage is that it is no longer tied to the details of
the dispatch semantics of the language being interpreted, but applies in
more general situations.  Promotion is thus the central enabling
primitive to make partial evaluation a practical approach to language
independent dynamic compiler generation.

XXX

\subsection{Promotion in Practise}

There are values that, if known at compile time, allow the JIT compiler to
produce very efficient code.  Unfortunately, these values are tipically red,
e.g. the exact type of a variable.

"Promotion" is a particular operation that convert a red value into a green
value; i.e., after the promotion of a variable, the JIT compiler knows its
value at compile time. Since the value of a red variable is not known until
runtime, we need to postpone the compilation phase after the runtime phase.

This is done by continuously intermixing compile time and runtime; a promotion
is implemented in this way:

@@

@@

Interpreters for dynamic languages typically allocate a lot of small
objects, for example due to boxing.  For this reason, we
implemented a way for the compiler to generate residual memory
allocations as lazily as possible.  The idea is to try to keep new
run-time structures "exploded": instead of a single run-time pointer to
a heap-allocated data structure, the structure is "virtualized" as a set
of fresh variables, one per field.  In the compiler, the variable that
would normally contain the pointer to the structure gets instead a
content that is neither a run-time value nor a compile-time constant,
but a special \emph{virtual structure} – a compile-time data structure that
recursively contains new variables, each of which can again store a
run-time, a compile-time, or a virtual structure value.

This approach is based on the fact that the "run-time values" carried
around by the compiler really represent run-time locations – the name of
a CPU register or a position in the machine stack frame.  This is the
case for both regular variables and the fields of virtual structures.
It means that the compilation of a \texttt{getfield} or \texttt{setfield}
operation performed on a virtual structure simply loads or stores such a
location reference into the virtual structure; the actual value is not
copied around at run-time.

@@

Virtual structures still avoid the run-time allocation of most
short-lived objects, even in non-trivial situations.  The following
example shows a typical case.  Consider the Python expression \texttt{a+b+c}.
Assume that \texttt{a} contains an integer.  The PyPy Python interpreter
implements application-level integers as boxes – instances of a
\texttt{W\_IntObject} class with a single \texttt{intval} field.  Here is the
addition of two integers:

@@

When interpreting the bytecode for \texttt{a+b+c}, two calls to \texttt{add()} are
issued; the intermediate \texttt{W\_IntObject} instance is built by the first
call and thrown away after the second call.  By contrast, when the
interpreter is turned into a compiler, the construction of the
\texttt{W\_IntObject} object leads to a virtual structure whose \texttt{intval}
field directly references the register in which the run-time addition
put its result.  This location is read out of the virtual structure at
the beginning of the second \texttt{add()}, and the second run-time addition
directly operates on the same register.

>>>>

In the sequel, we describe in more details one of the main new
techniques introduced in our approach, which we call \emph{promotion}.  In
short, it allows an arbitrary run-time (i.e. red) value to be turned into a
compile-time (i.e. green) value at any point in time.  Promotion is thus the central way by
which we make use of the fact that the JIT is running interleaved with actual
program execution. Each promotion point is explicitly defined with a hint that
must be put in the source code of the interpreter.

@@

Promotion requires interleaving compile-time and run-time phases,
otherwise the compiler can only use information that is known ahead of
time. It is impossible in the "classical" approaches to partial
evaluation, in which the compiler always runs fully ahead of execution
This is a problem in many realistic use cases.  For example, in an
interpreter for a dynamic language, there is mostly no information
that can be clearly and statically used by the compiler before any
code has run.

A very different point of view on promotion is as a generalization of techniques
that already exist in dynamic compilers as found in modern virtual machines for
object-oriented language.  In this context feedback
techniques are crucial for good results.  The main goal is to
optimize and reduce the overhead of dynamic dispatching and indirect
invocation.  This is achieved with variations on the technique of
polymorphic inline caches \cite{XXX}: the dynamic lookups are cached and
the corresponding generated machine code contains chains of
compare-and-jump instructions which are modified at run-time.  These
techniques also allow the gathering of information to direct inlining for even
better optimization results.

In the presence of promotion, dispatch optimization can usually be
reframed as a partial evaluation task.  Indeed, if the type of the
object being dispatched to is known at compile-time, the lookup can be
folded, and only a (possibly even inlined) direct call remains in the
generated code.  In the case where the type of the object is not known
at compile-time, it can first be read at run-time out of the object and
promoted to compile-time.  As we will see in the sequel, this produces
machine code very similar to that of polymorphic inline
caches\footnote{Promotion can also be seen as a generalization of
a partial evaluation transformation called "The Trick" (see e.g. \cite{XXX}),
which again produces similar code but which is only applicable for finite sets
of values.}.

The essential advantage of promotion is that it is no longer tied to the details of
the dispatch semantics of the language being interpreted, but applies in
more general situations.  Promotion is thus the central enabling
primitive to make partial evaluation a practical approach to language
independent dynamic compiler generation.

\subsection{Implementing Promotion}

The implementation of promotion requires a tight coupling between
compile-time and run-time: a \emph{callback}, put in the generated code,
which can invoke the compiler again.  When the callback is actually
reached at run-time, and only then, the compiler resumes and uses the
knowledge of the actual run-time value to generate more code.

The new generated code is potentially different for each run-time value
seen.  This implies that the generated code needs to contain some sort
of updatable switch, which can pick the right code path based on the
run-time value.

\cfbolz{I think this example is confusing, it is unclear in which order things
happen how. I will try to come up with something different}.

@@

XXX maybe a tlc example can be used for the example above?

@@

Interpreters for dynamic languages typically continuously allocate a lot of small
objects, for example due to boxing. This makes arithmetic operations extremely
inefficient. For this reason, we
implemented a way for the compiler to try to avoid memory allocations in the
residual code as long as possible. The idea is to try to keep new
run-time structures "exploded": instead of a single run-time object allocated on
the heap, the object is "virtualized" as a set
of fresh variables, one per field. Only when the object can be accessed by from
somewhere else is it actually allocated on the heap. The effect of this is similar to that of
escape analysis \cite{XXX}, which also prevents allocations of objects that can
be proven to not escape a method or set of methods.

@@

Virtual structures still avoid the run-time allocation of most
short-lived objects, even in non-trivial situations.  The following
example shows a typical case.  XXX use TLC example

The PyPy Python interpreter
implements application-level integers as boxes – instances of a
\texttt{W\_IntObject} class with a single \texttt{intval} field.  Here is the
addition of two integers:

@@

XXX kill the rest?!‽
When interpreting the bytecode for \texttt{a+b+c}, two calls to \texttt{add()} are
issued; the intermediate \texttt{W\_IntObject} instance is built by the first
call and thrown away after the second call.  By contrast, when the
interpreter is turned into a compiler, the construction of the
\texttt{W\_IntObject} object leads to a virtual structure whose \texttt{intval}
field directly references the register in which the run-time addition
put its result.  This location is read out of the virtual structure at
the beginning of the second \texttt{add()}, and the second run-time addition
directly operates on the same register.
